Generation of TypeScript Declaration Files from JavaScript Code

### 2020-06-26 ecoop-2020/main.tex
Add new evaluation concept and fix different todos

This work presents \texttt{dts-generate}, a tool that generates
TypeScript declaration files for JavaScript libraries uploaded to the \NPM{}
registry. It extracts code examples from the documentation written by
the developer, executes the library driven by the examples, gathers
run-time information, and generates a declaration file based on this
information. To evaluate the tool, 244 declaration files were generated and 
compared with the declaration file provided on DefinitelyTyped.
33 files out of 244 had no differences.
% but what about the remaining generated files
\end{abstract}

@@

\begin{lstlisting}[
    caption={\textbf{Unexpected JavaScript behavior} - falsy values, \lstinline!typeof!, \lstinline!null! and \lstinline!undefined! operators and type coercion},
    language=JavaScript,
	label=code:introduction-javascript-wtfs,
    float,
    abovecaptionskip=-\medskipamount
]
"0" == false; // true
true == "1.00"; // true
false == "    \n\r\t     "; // true
false == []; // true
0 == []; // true
null == undefined; // true
[1] + 1; //'11'
[2] == "2"; // true
null + undefined + [1, 2, 3] // 'NaN1,2,3'
"hello world".lenth + 1 // NaN | note 'lenth' instead of 'length'
[1, 15, 20, 100].sort() // [ 1, 100, 15, 20 ]
typeof null; // object
null instanceof Object; // false
\end{lstlisting}
% pjt: why unexpected?
% "01" < "00100"; // false

@@

\section{Motivating Example}
\label{sec:motivating-example}
The \NPM{} package \texttt{route-parser} is a simple route parsing,
matching, and reversing library for
Javascript\footnote{https://www.npmjs.com/package/route-parser}. It
has about 35,000 weekly downloads and 221 \NPM{} packages depend on it. If
a developer creates or extends TypeScript code that depends on the
\texttt{route-parser} library, the TypeScript
compiler and IDE requires a declaration file for that library to
perform static checking and code completion, respectively. We 
use \texttt{dts-generate} to automatically generate a TypeScript
declaration file for \texttt{route-parser}. The tool downloads the \NPM{} 
package, runs the examples extracted from its documentation, gathers
run-time information, and generates a TypeScript declaration
file. The result is 
shown in \figref{fig:motivating-example-route-parser-vscode} and it is
ready for use in a TypeScript project. For example,  Visual Studio's
code completion runs properly with it
(\figref{fig:motivating-example-route-parser-vscode}). If the
\texttt{route-parser} package gets modified in the future, a new declaration
file can be generated automatically using
\texttt{dts-generate}. Our comparator tool can compare the new file
for incompatibilities with the previous declaration file.

\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.70\linewidth}
    \begin{lstlisting}[language=TypeScript]
export = Route;

declare class Route {
  constructor(spec: string);
  match(path: string): object;
  reverse(params: object): string;
}

declare namespace Route {
}
    \end{lstlisting}
  \end{subfigure}

@@

  \begin{subfigure}{1.\linewidth}
    \centering
    \includegraphics[width=0.7\linewidth]{motivating-example-route-parser-vscode.png}
  \end{subfigure}

  \caption{\textbf{Declaration file for \texttt{route-parser} generated with \texttt{dts-generate}} - Constructor and methods are correctly identified. Declaration file can be correctly used in Visual Studio Code\footnote{https://code.visualstudio.com}.}
  \label{fig:motivating-example-route-parser-vscode}
\end{figure}

@@

The declaration file shown in
\figref{fig:motivating-example-route-parser-vscode} describes a
package with a single exported class \lstinline{Route}. The
description of the class comprises the type signature of the
constructor (line~4) and the methods \lstinline{match} and
\lstinline{reverse} (lines~5 and~6).

!!!TODO: explain meaning of namespace.

@@

% $ ./dts-generate abs
% $ cat output/abs/index.d.ts 
\begin{lstlisting}[
    caption={Example for template \textbf{module-function}},
    language=TypeScript,
	label=code:dts-generate-example,
    float,
    abovecaptionskip=-\medskipamount
]
export = Abs;

declare function Abs(input: string): string;
\end{lstlisting}
!!!TODO: example for simple \textbf{module}

The \texttt{route-parser} library is an instance of a
\textbf{module-class}. The \texttt{abs} library provides a single
function \lstinline{Abs} so it uses template \textbf{module-function}
as shown in \ref{code:dts-generate-example}.

@@

We considered option~\ref{item:2} under the assumption that most
libraries would come with test cases. However, there is no standard
for testing JavaScript code so that test cases were difficult to reap
from the \NPM{} packages: they use different directory structures,
employ differing (or no) testing tools, or do not have tests
at all.

In the end, option~\ref{item:3} was the most viable even though there
is no standard for documentation, either. However, almost all repositories
contain README files where the library authors briefly describe in prose what
the code does, which problem it solves, how to install the
application, how to build the code, etc. It is very common that
developers provide code examples in the README files to show how the
library works and how to use it. This observation holds in particular
for \NPM{} packages, which are generally created to solve a specific
problem of JavaScript development.

@@

For example, in the case of \texttt{route-parser} ...

@@

The implementation stores these observations as entities called
\texttt{interactions}. They are used for translating, modifying, and
aggregating Jalangi's raw event information to get an application
specific data representation. The run-time information is saved as a
JSON file that can be used for later processing. The tool is written
in JavaScript and runs in \NodeJS{} in a Docker container.

@@

!!!TODO too vague: more information --- how do you choose between the templates?

We use different fields (which?) from the run-time information to detect
how the module is used to choose the most suitable template. The implemented
templates are \textbf{module}, \textbf{module-class} and
\textbf{module-function}.

@@

!!!TODO: insufficient explanation of \texttt{followingInteractions}
--- needs to be mentioned in the previous step

\section{Evaluation}
\label{sec:dts-generate-evaluation}
After generating a declaration file for an \NPM{} package, we need to
evaluate its quality. To this end, we compare the generated
declaration file against the one uploaded to the DefinitelyTyped
repository for the same module, as shown in
\figref{fig:evaluation-diagram}. While this approach does not provide
an absolute measure of quality, it gives us at least an indication of
the pragmatics of \texttt{dts-generate}: The files on DefinitelyTyped
are perceived to be useful by the community. If the accuracy of the
generated files is comparable with the accuracy of the files on
DefinitelyTyped, then \texttt{dts-generate} is a viable alternative to
manually creating a definition file.

!!!TODO: check how many files on DT are just generated by
\texttt{dts-gen} (e.g., the TypeScript tool)

@@

To this end, we need a means of comparing two declaration files. Of
course, we are not interested in a textual comparison, but in a
comparison of the structures described by the files. Technically, our
comparison of two structures A and B yields one of five outcomes:
\begin{itemize}
\item A and B are equivalent;
\item A is more general than B;
\item B is more general than A;
\item A and B have a common generalization (i.e., they are
  compatible);
\item none of the above holds (i.e., A and B are incompatible).
\end{itemize}

The first step is to parse declaration files, which can be
achieved by using the TypeScript Compiler API, a library traverse the
Abstract Syntax Tree of a TypeScript program
\cite{typescript-compiler-api}. 
The step also performs a sanity check of the generated declaration
files as it rejects files with syntactic or semantics errors.
The output of the parser is a structure where declared
{interfaces}, {functions}, {classes} and
{namespaces} are stored separately. Function arguments are
correctly described, identifying complex types like union types or
callbacks. Optional parameters are also identified. For
{classes}, a distinction between the constructor and methods
is made.

% The tool is called \texttt{parse-dts} and is naturally written in
% TypeScript. It also runs in \NodeJS{} in its corresponding Docker
% container.

It remains to compare the parsed representations of two declaration
files.
% It was discovered that the implementation was easier when
% focusing on each template independently.
As an example, we discuss the implementation of comparison for
the \textbf{module-function} template.

The following criteria were applied:
\begin{itemize}
\item Number of declared functions: Checks the number of declared
  and exported functions for each of both declaration files. 
\item Name of declared functions: Checks whether both of the
  declaration files declared a function with the same name. 
\item Number of parameters: Checks the number of parameters of the
  declared functions. This is checked for optional and non-optional
  parameters.

  !!!TODO: what about parameter names? what about object
  types/interface types of parameters?
\item Interfaces: Checks the number of declared interfaces and the
  fields within those interfaces. 
\item Errors: Indicates whether there are errors in the declaration files.
\end{itemize}

!!!TODO: too informal. we need to define an ordering on interfaces etc.

\section{Results}
\label{sec:results}

!!!TODO: first we need to say, what we want to achieve with the
experimental evaluation

@@

Figure \ref{fig:experiments-overall-funnel} shows that a declaration
file was generated for 244 modules out of 6029 modules. Samples of the
generated declaration files for templates \textbf{module},
\textbf{module-class} and \textbf{module-function} are presented in
\secref{sec:experiments-declaration-files-generation}.

@@

!!!TODO: the remaining two stages ``run-time information'' and
``generated declaration files'' need to be explained

@@

\figref{fig:experiments-results-module-function} shows the generated
declaration files for simple modules like \texttt{abs},
\texttt{dirname-regex}, and \texttt{escape-html}. All of them were
generated using the \textbf{module-function} template. There are no
differences between the generated files and the corresponding
declaration files on DefinitelyTyped.

The left side of the figure shows the generated declaration file with
\lstinline{dts-generate}, the right side shows the corresponding file
in the DefinitelyTyped repository. Functions are correctly detected; input
and output types are accurately inferred.

!!!TODO: explain, what about
\begin{itemize}
\item uppercase/lowercase in dirname-regex?
\item extra namespace in escape-html?
\end{itemize}

@@

Templates of type \textbf{module-class} are shown for modules
\texttt{flake-idgen}, \texttt{route-parser}, and \texttt{timer-machine}
in \figref{fig:experiments-results-module-class-flake-idgen} and
\figref{fig:experiments-results-module-class-timer-machine},
respectively. Properties of interfaces and class methods are correctly
generated. Optional parameters are not detected, as it was not
considered for the implementation.

For \texttt{flake-idgen},  the parameters of interface
\texttt{ConstructorOptions} are correctly detected. The name of the
interface differs because it is automatically generated based on the
name of the argument variable. Optional properties were not
implemented, which explains the type \lstinline{undefined} for some
properties. Analogously, the callback \texttt{cb} is inferred as
\lstinline{undefined}.

!!!TODO:
\begin{itemize}
\item why is the callback undefined? is it due to the example? (what does it
  look like?) why is it called \texttt{cb}?
\item the export / class is named differently; doesn't that matter?
\end{itemize}

For \texttt{timer-machine},  the parameter \lstinline{started} is
inferred as \lstinline{undefined} instead of marking it as
optional. Methods that were not executed cannot appear in the
generated declaration file. For that reason, a type corresponding to
\lstinline{TimerEvent} is not generated, either.

\begin{figure}[tp]
    \centering
    \begin{subfigure}{0.48\linewidth}
      \begin{lstlisting}[language=TypeScript]
export = FlakeIdgen;

declare class FlakeIdgen {
    constructor(options: FlakeIdgen.I__options);
    next(cb: undefined): Buffer;
}

>>>>

This work presents \texttt{dts-generate}, a tool that generates
TypeScript declaration files for JavaScript libraries uploaded to the \NPM{}
registry. It extracts code examples from the documentation written by
the developer, executes the library driven by the examples, gathers
run-time information, and generates a declaration file based on this
information. To evaluate the tool, 244 declaration files were generated and 82 were
compared with the corresponding declaration file provided on DefinitelyTyped. 72 files presented positive results, having no differences at all or differences that can be solved by modifying the code examples accordingly.

\end{abstract}

@@

\begin{lstlisting}[
    caption={\textbf{Unintuitive JavaScript behavior} - falsy values, \lstinline!typeof!, \lstinline!null! and \lstinline!undefined! operators and type coercion},
    language=JavaScript,
	label=code:introduction-javascript-wtfs,
    float,
    abovecaptionskip=-\medskipamount
]
"0" == false; // true
true == "1.00"; // true
false == "    \n\r\t     "; // true
false == []; // true
0 == []; // true
null == undefined; // true
[1] + 1; //'11'
[2] == "2"; // true
null + undefined + [1, 2, 3] // 'NaN1,2,3'
"hello world".lenth + 1 // NaN | note 'lenth' instead of 'length'
[1, 15, 20, 100].sort() // [ 1, 100, 15, 20 ]
typeof null; // object
null instanceof Object; // false
"01" < "00100"; // false
\end{lstlisting}
% pjt: why unexpected?

@@

\section{Motivating Example}
\label{sec:motivating-example}
The \NPM{} package \texttt{glob-to-regexp} is a simple JavaScript library, which turns a glob expression into a regular expression\footnote{https://www.npmjs.com/package/glob-to-regexp}. It
has about 6,500,000 weekly downloads and 181 \NPM{} packages depend on it. If
a developer creates or extends TypeScript code that depends on the
\texttt{glob-to-regexp} library, the TypeScript
compiler and IDE requires a declaration file for that library to
perform static checking and code completion, respectively. We 
use \texttt{dts-generate} to automatically generate a TypeScript
declaration file for \texttt{glob-to-regexp}. The tool downloads the \NPM{} 
package, runs the examples extracted from its documentation, gathers
run-time information, and generates a TypeScript declaration
file. The result is 
shown in \figref{fig:motivating-example-glob-to-regexp-vscode} and it is
ready for use in a TypeScript project. For example,  Visual Studio's
code completion runs properly with it
(\figref{fig:motivating-example-glob-to-regexp-vscode}). If the
\texttt{glob-to-regexp} package gets modified in the future, a new declaration
file can be generated automatically using
\texttt{dts-generate}. Our comparator tool can compare the new file
for incompatibilities with the previous declaration file.

\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.70\linewidth}
    \begin{lstlisting}[language=TypeScript]
export = GlobToRegexp;

declare function GlobToRegexp(glob: string, opts: undefined): RegExp;
declare function GlobToRegexp(glob: string, opts: GlobToRegexp.I__opts): RegExp;
declare namespace GlobToRegexp {
  export interface I__opts {
    'extended': undefined | boolean;
    'globstar': undefined | boolean;
    'flags': undefined | string;
  }

}
    \end{lstlisting}
  \end{subfigure}

@@

  \begin{subfigure}{1.\linewidth}
    \centering
    \includegraphics[width=0.8\linewidth]{motivating-example-glob-to-regexp-vscode.png}
  \end{subfigure}

  \caption{\textbf{Declaration file for \texttt{glob-to-regexp} generated with \texttt{dts-generate}} - Interfaces is correctly detected. Optional parameters are explicitly declared as \texttt{undefined}. Declaration file can be correctly used in Visual Studio Code\footnote{https://code.visualstudio.com}.}
  \label{fig:motivating-example-glob-to-regexp-vscode}
\end{figure}

@@

The declaration file shown in
\figref{fig:motivating-example-glob-to-regexp-vscode} describes a
package with a single exported function. The first parameter is of type \texttt{string} and the second one is an optional interface. TypeScript namespaces are used for organizing types declared in a declaration file and avoiding name collisions with other types \cite{typescript-namespaces}. For example, the interface \texttt{I\_\_opts} declared in \figref{fig:motivating-example-glob-to-regexp-vscode} is used as \lstinline[language=TypeScript]{globToRegexp.I__opts}, since it is declared within a namespace.

@@

\begin{lstlisting}[
    caption={Example for template \textbf{module-function}},
    language=bash,
	label=code:dts-generate-example,
    float,
    abovecaptionskip=-\medskipamount
]
$ ./dts-generate abs
$ cat output/abs/index.d.ts 
export = Abs;

declare function Abs(input: string): string;
\end{lstlisting}

TypeScript provides a guide on how to write high quality declaration files\footnote{https://www.typescriptlang.org/docs/handbook/declaration-files/by-example.html}. They explain the main concepts through examples. We extracted one example for each template and generated used \texttt{dts-generate} to generate a corresponding declaration file, as presented in \figref{fig:typescript-templates-by-example}. The \texttt{glob-to-regexp} library is an instance of a \textbf{module-function}. It is worth mentioning that templates are a way of describing how a JavaScript library is intended to be used. Although highly unlikely, it is possible to create a library that can be represented with a different template depending on how it is used, as shown in \figref{fig:multi-purpose-module-example}. Run-time analysis proves useful in this case, since it is not sufficient to statically analyze the exported module.

\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.48\linewidth}
    \begin{lstlisting}[language=JavaScript]
var greet = require("./greet-settings-module");

greet({
  greeting: "hello world",
  duration: 4000
});

greet({
  greeting: "hello world",
  color: "#00ff00"
});
    \end{lstlisting}
    \caption{TypeScript example for \texttt{module-function} template}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.48\linewidth}
    \begin{lstlisting}[language=TypeScript]
export = GreetSettingsModule;

declare function GreetSettingsModule(settings: GreetSettingsModule.I__settings): void;
declare namespace GreetSettingsModule {
  export interface I__settings {
    'greeting': string;
    'duration': number | undefined;
    'color': undefined | string;
  }

}
    \end{lstlisting}
    \caption{Declaration file for \texttt{module-function} template}
  \end{subfigure}

  \begin{subfigure}{0.48\linewidth}
      \begin{lstlisting}[language=JavaScript]
var myLib = require("./greet-module");

var result = myLib.makeGreeting("hello, world");
console.log("The computed greeting is: " + result);

var goodbye = myLib.makeGoodBye();
console.log("The computed goodbye is: " + goodbye);    
      \end{lstlisting}
      \caption{Example for \texttt{module}}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.48\linewidth}
      \begin{lstlisting}[language=TypeScript]
export function makeGreeting(str: string): string;
export function makeGoodBye(): string;        
      \end{lstlisting}
      \caption{Declaration file for \texttt{module} template}
    \end{subfigure}

    \begin{subfigure}{0.48\linewidth}
      \begin{lstlisting}[language=JavaScript]
var Greeter = require("./greet-classes-module.js");

var myGreeter = new Greeter("hello, world");
myGreeter.greeting = "howdy";
myGreeter.showGreeting();
      \end{lstlisting}
      \caption{Example for \texttt{module-class}}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.48\linewidth}
      \begin{lstlisting}[language=TypeScript]
export = Greeter;

declare class Greeter {
  constructor(message: string);
  showGreeting(): void;
}

declare namespace Greeter {
}
      \end{lstlisting}
      \caption{Declaration file for \texttt{module-class} template}
    \end{subfigure}

  \caption{Results for \textbf{module-function} template}
  \label{fig:typescript-templates-by-example}
\end{figure}

\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.80\linewidth}
      \begin{lstlisting}[language=JavaScript]
var multiPurposeModule = require('./multi-purpose-module');

// module-function
console.log(multiPurposeModule("John", "Doe"));
// John Doe

// module-class
var m = new multiPurposeModule("Jane", "Doe");
console.log(m.firstName + " - " + m.lastName);
// Jane - Doe

console.log(m.sayHello());
// Hello, my name is Jane Doe

// module
console.log(multiPurposeModule.anotherFunction());
// I am another function! 
      \end{lstlisting}
      \caption{index.js}
    \end{subfigure}

    \begin{subfigure}{0.80\linewidth}
      \begin{lstlisting}[language=TypeScript]
var multiPurposeModule = function(firstName, lastName) {
  this.firstName = firstName;
  this.lastName = lastName;

  return firstName + " " + lastName;
};

multiPurposeModule.prototype.sayHello = function() {
  return "Hello, my name is " + this.firstName + " " + this.lastName;
};

multiPurposeModule.anotherFunction = function() {
  return "I am another function!";
};

module.exports = multiPurposeModule;   
      \end{lstlisting}
      \caption{multi-purpose-module/index.js}
    \end{subfigure}
  \caption{Example module that implements all three templates simultaneously}
  \label{fig:multi-purpose-module-example}
\end{figure}

@@

We considered option~\ref{item:2} under the assumption that most
libraries would come with test cases. However, there is no standard
for testing JavaScript code so that test cases were difficult to reap
from the \NPM{} packages: they use different directory structures,
employ differing (or no) testing tools, or do not have tests
at all. Moreover, developers try to cover edge cases or even not allowed values in their tests. Such tests are not really useful for describing a library from the consumer's point of view. For example, a developer might write a test invoking a function with \texttt{null} just to validate that it throws an error in such scenario.

In the end, option~\ref{item:3} was the most viable even though there
is no standard for documentation, either. However, almost all repositories
contain README files where the library authors briefly describe in prose what
the code does, which problem it solves, how to install the
application, how to build the code, etc. It is very common that
developers provide code examples in the README files to show how the
library works and how to use it. This observation holds in particular
for \NPM{} packages, which are generally created to solve a specific
problem of JavaScript development. These code fragments do not present the problem of option~\ref{item:2} since they are meant to be used by developers using the package.

@@

The extracted example for the case of \texttt{glob-to-regexp} is shown in \coderef{code:code-example-extracted-motivating-example}.

\begin{lstlisting}[
  caption={Extracted example for \texttt{glob-to-regexp}},
  language=JavaScript,
label=code:code-example-extracted-motivating-example,
  float,
  abovecaptionskip=-\medskipamount
]
var globToRegExp = require('glob-to-regexp');
var re = globToRegExp("p*uck");
re.test("pot luck"); // true
re.test("pluck"); // true
re.test("puck"); // true

re = globToRegExp("*.min.js");
re.test("http://example.com/jquery.min.js"); // true
re.test("http://example.com/jquery.min.js.map"); // false

re = globToRegExp("*/www/*.js");
re.test("http://example.com/www/app.js"); // true
re.test("http://example.com/www/lib/factory-proxy-model-observer.js"); // true

// Extended globs
re = globToRegExp("*/www/{*.js,*.html}", { extended: true });
re.test("http://example.com/www/app.js"); // true
re.test("http://example.com/www/index.html"); // true
\end{lstlisting}

@@

The implementation stores these observations as entities called
\texttt{interactions}. They are used for translating, modifying, and
aggregating Jalangi's raw event information to get an application
specific data representation. Each function invocation is stored as a \texttt{FunctionContainer} which contains an \texttt{ArgumentContainer} for each argument. Each \texttt{ArgumentContainer} contains the name and index of the argument and a collection of \texttt{interactions}. The interaction \texttt{getField} gets triggered whenever a property of an object gets accessed and it tracks the name of the accessed field. The invocation of an object's method is tracked with the \texttt{methodCall} interaction. The \texttt{usedAsArgument} interaction gets triggered when an argument is used in the invocation of a function. The \texttt{followingInteractions} property is used for inferring nested interfaces. It is an array that recursively records \texttt{interactions} on the return value of \texttt{getField} or \texttt{methodCall}.

We wrap each function's argument around a wrapper object, which enables to store meta-information and greatly simplifies the mapping of an observation with the corresponding \texttt{ArgumentContainer} or \texttt{interaction}. We then use the original values for critical operators such as \texttt{===} or \texttt{typeof}, for which wrapper objects will definitely modify the behavior of the code.

The property \texttt{requiredModule} stores the name of the module that declared the invoked function. If a function is explicitly exported by the module, the property \texttt{isExported} will be set to \texttt{true} when it is invoked. If a function of an exported object is invoked, \texttt{isExported} will be \texttt{false} and \texttt{requiredModule} will contain the name of the required module.

The run-time information is saved as a JSON file that can be used for later processing. The tool is written in JavaScript and runs in \NodeJS{} in a Docker container.

@@

Instead of analyzing the shape of the exported module, we inspect how the module is used to choose between the templates. We analyze the properties \texttt{isExported}, \texttt{requiredModule} and \texttt{isConstructor} from the run-time JSON to distinguish between \texttt{module-class} and \texttt{module-function}. If a function is invoked and it is imported from the module we are analyzing we infer the template \texttt{module-class} or \texttt{module-function}. We choose \texttt{module-class} if the function is used as a constructor. If not, we choose \texttt{module-function}. Otherwise, we infer the \texttt{module} template. \figref{fig:dts-generate-choose-templates} exposes this logic using the example provided in \figref{fig:multi-purpose-module-example}.

\begin{figure}[tp]
  \centering
  \begin{subfigure}{0.48\linewidth}
    \begin{lstlisting}[language=JavaScript]
var multiPurposeModule = require('./multi-purpose-module');

multiPurposeModule("John", "Doe");
// John Doe
    \end{lstlisting}
    \caption{Example module \texttt{multi-purpose-module} used as \texttt{module-function}}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.48\linewidth}
    \begin{lstlisting}[language=TypeScript]
export = MultiPurposeModule;

declare function MultiPurposeModule(firstName: string, lastName: string): string;
    \end{lstlisting}
    \caption{Generated declaration file with \texttt{module-function} template}
  \end{subfigure}

  \begin{subfigure}{0.48\linewidth}
    \begin{lstlisting}[language=JavaScript]
var multiPurposeModule = require('./multi-purpose-module');

var m = new multiPurposeModule("Jane", "Doe");
var fullName = m.firstName + " - " + m.lastName;
// Jane - Doe

m.sayHello();
// Hello, my name is Jane Doe
    \end{lstlisting}
    \caption{Example module \texttt{multi-purpose-module} used as \texttt{module-class}}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.48\linewidth}
      \begin{lstlisting}[language=TypeScript]
export = MultiPurposeModule;

declare class MultiPurposeModule {
  constructor(firstName: string, lastName: string);
  sayHello(): string;
}

declare namespace MultiPurposeModule {
}
      \end{lstlisting}
      \caption{Generated declaration file with \texttt{module-class} template}
    \end{subfigure}

    \begin{subfigure}{0.48\linewidth}
      \begin{lstlisting}[language=JavaScript]
var multiPurposeModule = require('./multi-purpose-module');

multiPurposeModule.anotherFunction();
// I am another function!
      \end{lstlisting}
      \caption{Example module \texttt{multi-purpose-module} used as \texttt{module}}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.48\linewidth}
      \begin{lstlisting}[language=TypeScript]
export function anotherFunction(): string;
      \end{lstlisting}
      \caption{Generated declaration file with \texttt{module} template}
    \end{subfigure}

  \caption{Different ways of using example module presented in \figref{fig:multi-purpose-module-example}. \texttt{dts-generate} generates infers the correct template for each case. The JavaScript implementation of the module remained unmodified for each example.}
  \label{fig:dts-generate-choose-templates}
\end{figure}

We implemented only basic TypeScript features and we focused our effort in building and end-to-end solution that generates valid and useful declaration files. Only the basic types \texttt{string}, \texttt{number}, \texttt{boolean} and its union were considered both for function parameters and interface properties. Common types such as \texttt{any}, arrays, function callbacks, tuples, intersections and features such as generics, index signatures were not implemented. Each declaration file was tagged with \texttt{dts-parse} and those containing unimplemented features were ignored. We add the \texttt{undefined} type for describing an optional parameter instead of adding the corresponding token \texttt{?}, which is also done by TypeScript in strict null checking mode\cite{typescript-optional-parameters-and-properties}.

@@

\section{Evaluation}
\label{sec:dts-generate-evaluation}
After generating a declaration file for an \NPM{} package, we need to
evaluate its quality. To this end, we created two tools:
\begin{itemize}
  \item \texttt{dts-parse}: A TypeScript declaration files parser, which transforms the file into a JSON file with a defined structure.
  \item \texttt{dts-compare}: A tool that uses \texttt{dts-parse} and compares two TypeScript declaration files computing differences by applying several criteria.
\end{itemize}

We applied \texttt{dts-compare} to compare the generated declaration file against the one uploaded to the DefinitelyTyped repository for the same module, as shown in \figref{fig:evaluation-diagram}. While this approach does not provide an absolute measure of quality, it gives us at least an indication of the pragmatics of \texttt{dts-generate}: The files on DefinitelyTyped
are perceived to be useful by the community. If the accuracy of the
generated files is comparable with the accuracy of the files on
DefinitelyTyped, then \texttt{dts-generate} is a viable alternative to
manually creating a definition file.

%!!!TODO: check how many files on DT are just generated by \texttt{dts-gen} (e.g., the TypeScript tool)

@@

\subsection{dts-parse}
We need a means of comparing two declaration files. Of
course, we are not interested in a textual comparison, but in a
comparison of the structures described by the files.

The first step is to parse the declaration files, which can be
achieved by using the TypeScript Compiler API, a library which traverse the
Abstract Syntax Tree of a TypeScript program
\cite{typescript-compiler-api}. 
The step also performs a sanity check of the generated declaration
files as it rejects files with syntactic or semantics errors.
The output of \texttt{dts-parse} is a structure where declared
{interfaces}, {functions}, {classes} and
{namespaces} are stored separately. Function arguments are
correctly described, identifying complex types like union types or
callbacks. Optional parameters are also identified. For
{classes}, a distinction between the constructor and methods
is made. Declaration files are tagged according to their features, which is useful for identifying and filtering out declaration files containing unimplemented features.

\texttt{dts-parse} is itself written in TypeScript. It receives the file as input parameter and returns a JSON, as shown in \coderef{code:dts-parse-example}.

\begin{lstlisting}[
  caption={Example for \texttt{dts-parse} executed on module \texttt{glob-to-regexp} and combined with \texttt{jq} to browse through the returned JSON},
  language=bash,
label=code:dts-parse-example,
  float,
  abovecaptionskip=-\medskipamount
]
$ dts-parse -i glob-to-regexp/index.d.ts | jq '.parsing.functions[0].name'
"GlobToRegexp"
\end{lstlisting}

\subsection{dts-compare}
\label{sec:dts-compare}
The tool is written in TypeScript and receives 2 declaration files as an input and returns a JSON containing the result of the comparison, as shown in \coderef{code:dts-compare-example}. Following the naming convention of any standard Unit Testing framework, the input files are flagged as \lstinline{expected} or \lstinline{actual}. We assigned the \lstinline{expected} flag to the DefinitelyTyped declaration file.

\begin{lstlisting}[
  caption={Example for \texttt{dts-compare} executed on declaration files without differences},
  language=bash,
label=code:dts-compare-example,
  float,
  abovecaptionskip=-\medskipamount
]
$ dts-compare -e expected.d.ts -a actual.d.ts --module-name "my-module"
{
    "module": "my-module",
    "template": "module",
    "differences": [],
    "tags": []
}
\end{lstlisting}

\texttt{dts-compare} applies \texttt{dts-parse} internally to both files to get a common structure eligible for comparison. The result of the comparison is an array of an abstract entity called \texttt{Difference}. We extracted code examples written by developers to execute the code and gather run-time information. The quality of the generated declaration file is heavily dependent on the code examples. For example, if a property of an interface does not get accessed due to a code example not exploring a particular execution path, that property will not exist in the generated declaration file. Consequently, in order to isolate the impact of incomplete code examples on our quality analysis we identified each difference as follows:
\begin{itemize}
  \item \texttt{solvable}: Differences that can be solved by expanding the code examples. For example, invoking a function with different parameters so that a missing interface property gets accessed.
  \item \texttt{unsolvable}: Differences that can not be solved are marked as \texttt{unsolvable}. These differences are implementation errors of \texttt{dts-generate}.
\end{itemize}

The concrete evaluated differences are as follows:
\begin{itemize}
  \item \texttt{TemplateDifference}: A difference between the formal TypeScript templates. For example, if the file in DefinitelyTyped is written with \texttt{module} and we generated a file using \texttt{module-function}. The comparison stops if the templates differ.
  \item \texttt{ExportAssignmentDifference}: An equality check between the export assignments of both files, that is in the expression \lstinline{export = XXX}.
  \item \texttt{FunctionMissingDifference}: A function is present in the DefinitelyTyped file but not in the generated one. This applies to functions and methods of classes and interfaces.
  \item \texttt{FunctionMissingDifference}: A function is present in the generated declaration file but not in DefinitelyTyped.
  \item \texttt{FunctionOverloadingDifference}: The number of declarations for the same functions is different. This is particularly common when developers declare a function multiple times with different single types instead of using union types.
  \item \texttt{ParameterMissingDifference}: A parameter of a function or a property of an interface is not present in the generated file.
  \item \texttt{ParameterExtraDifference}: A parameter of a function or a property of an interface is present in the generated file but not in the DefinitelyTyped file.
  \item \texttt{ParameterTypeDifference}: A parameter of a function or a property of an interface is generated with a different type than in the DefinitelyTyped file. Here we differentiate between \texttt{SolvableDifference} and \texttt{UnsolvableDifference}.
  \begin{itemize}
    \item \texttt{SolvableDifference}: A type difference that can be solved by writing better code examples. This includes a basic type that is converted to a union type, a function overloading or a parameter or property that is marked as optional.
    \item \texttt{UnsolvableDifference}: Anything not considered as \texttt{SolvableDifference}.
  \end{itemize}
\end{itemize}

Type aliases are invisible to \texttt{dts-compare} as only final types were considered: the declaration \lstinline{type T = string | number; declare function F(a: T);} is equivalent to \lstinline{declare function F(a: string | number);}. The same concept applies for literal interfaces. \texttt{dts-compare} does not contemplate differences in function return types, since interface inference of function return types is not covered in this version of \texttt{dts-generate}.

\section{Results}
\label{sec:results}
We analyzed the obtained results focusing on the following aspects:
\begin{itemize}
  \item The quality of the inferred types, interfaces and module structure using runtime information.
  \item The benefits of using code examples provided by developers as a first approximation to execute the libraries and avoid a cold start problem.
  \item The usability of the generated declaration file and whether \texttt{dts-generate} can be used in a proper development environment.
\end{itemize}

@@

Figure \ref{fig:experiments-overall-funnel} shows that a declaration
file was generated for 244 modules out of 6029 modules and we identified 80 modules that have only the features implemented by \texttt{dts-generate}. We obtained positive results for 69 of them. A detailed explanation of the overall quality of the generated files is provided in \secref{sec:experiments-evaluation}. Samples of the
generated declaration files for templates \textbf{module},
\textbf{module-class} and \textbf{module-function} are presented in
\secref{sec:experiments-declaration-files-generation}.

@@

\paragraph*{Run-time Information}
Run-time information was extracted for only 442 out of 1448 modules with working code examples. As explained, in order to extract the run-time information, the behavior of the code under analysis was explicitly modified by wrapping the arguments around Wrapper Objects. Furthermore, Jalangi’s instrumentation itself caused some executions to fail, since the modules contained JavaScript features that are not supported by Jalangi. As a result, run-time information could not be extracted for 1006 modules. An instrumentation without user defined analysis modules was not applied, so it was not possible to determine which modules were failing only because of Jalangi’s own limitations.

\paragraph*{Generated Declaration Files}
A declaration file was generated for 244 out of 442 modules. Despite the correct execution of instrumented code examples, 198 modules did not contain suitable run-time information for generating a declaration file. The extracted code examples for these modules did not execute the library itself and therefore the collected run-time information was not useful for generating a declaration file.

@@

\figref{fig:experiments-results-module-function} shows the generated
declaration files for simple modules like \texttt{abs},
\texttt{dirname-regex}, and \texttt{escape-html}. All of them were
generated using the \textbf{module-function} template. Left side of the figure shows the generated declaration file with
\lstinline{dts-generate}, the right side shows the corresponding file
in the DefinitelyTyped repository. There are no relevant
differences between the files. Functions are correctly detected; input and output types are accurately inferred.

Differences in the export assignment by modules \texttt{dirname-regex} and \texttt{escape-html} is not problematic since the exported module can receive a custom name given by the developer. \texttt{dts-generate} automatically generates the export assignment by transforming the module name into camel case form, following TypeScript guidelines. The parameter names are extracted from the variable names of the JavaScript code. The JavaScript concrete implementation of module \texttt{escape-html} declares the first parameter of the exported function with the name \texttt{string} and not \texttt{text}, as specified in DefinitelyTyped. A difference in the names of the function parameters will anyway not affect the code compilation in any way. Furthermore, the omitted namespace in module \texttt{escape-html} is irrelevant. It is actually not required, since there is nothing declared within the namespace.

@@
